\documentclass{ximera}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, fill=red!50]
\tikzstyle{io} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum height=1cm, text centered, fill=blue!30]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, text width=3cm, fill=orange!50]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, text width=2cm, fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\title{Loops}
\begin{document}
\begin{abstract}
We give an introduction to for and while loops.
\end{abstract}
\maketitle

\section{Loops}

Loops are useful whenever you have an algorithm that requires multiple uses of the same step. For example, say we wanted to compute the sum $1+2+\cdots+n$ for some integer $n>0$. We have seen in a previous section that this can be done using two variables, one that tracks the current value of the sum and another that tracks whether or not our task is finished (see the flowchart below).

\begin{center}
	\begin{tikzpicture}
    \node [startstop] at (0,0) (start) {Start};
    \node [io] at (0,-2) (io1) {Input $n$};
    \node [process] at (0,-4) (process1) {$sum:=0$\\$i:=1$};
    \node [decision] at (0,-7) (decision) {$i\leq n$};
    \node [io] at (-5,-7) (io2) {Output $sum$};
    \node [process] at (0,-10) (process2) {$sum:=sum+i$\\$i:=i+1$};
    \node [startstop] at (-5,-10) (stop) {Stop};

    \draw [arrow] (start) -- (io1);
    \draw [arrow] (io1) -- (process1);
    \draw [arrow] (process1) -- (decision);
    \draw [arrow] (decision) -- node[anchor=east] {yes} (process2);
    \draw [arrow] (decision) -- node[anchor=south] {no} (io2);
    \draw [arrow] (io2) -- (stop);
    \draw [arrow] (process2) -- (3,-10) |- (decision);
    \end{tikzpicture}
\end{center}

The structure of the flowchart above is very suggestive of why we use the terminology `loop'. We will now go over two types of loops: for loops and while loops.

\section{For Loops}

The Python syntax for a for loop is as follows:

\begin{verbatim}
==============================
for variable in iterable:
        code
==============================
\end{verbatim}

In the above code, \verb|for| indicates the start of a for loop, \verb|variable| is just a variable name of your choice, \verb|in| is a keyword, and \verb|iterable| is a Python object that can be iterated over, like a list or a string. This Python code almost reads as an English statement. It is saying, for each item in \verb|iterable|, whose value we assign to \verb|variable| each time, perform the instructions given by \verb|code|.

The next few examples illustrate how \verb|variable| takes on the values of \verb|iterable|.

\begin{sageCell}
for n in [0,1,2,3]:
        print(n)
\end{sageCell}

\begin{sageCell}
for c in 'MAT1630':
        print(c)
\end{sageCell}

Below is an example of how to compute $1+2+\cdots+10$ using a for loop.

\begin{sageCell}
n = 100
s = 0

for i in [1,2,3,4,5,6,7,8,9,10]:
        s = s + i

print(s)
\end{sageCell}

If we wanted to compute $1+2+\cdots+n$ for some large or yet unknown value of $n$, we can already see that the previous example won't cut it. We need some general way of creating useful lists of numbers. We will use the \verb|range| function to help us with this. The \verb|range| function is commonly used in for loops. The inputs for the \verb|range| function are similar to those used for list slicing. See the SageCells below for examples of how to use the \verb|range| function.

The range function with 1 argument.
\begin{sageCell}
for i in range(10):
     print(i)
\end{sageCell}

The range function with 2 arguments.
\begin{sageCell}
for j in range(2,8):
     print(j)
\end{sageCell}

The range function with 3 arguments.
\begin{sageCell}
for k in range(3,12,2):
     print(k)
\end{sageCell}

Using the \verb|range| function, we can now write a for loop that computes $1+2+\cdots+100$.

\begin{sageCell}
s = 0
for i in range(1,101):
        s = s + i
print(s)
\end{sageCell}

We can modify, improve, and generalize the loop above. We will change out the notation \verb|s = s + i| and use \verb|s += i| instead. (The \verb|+=| operator increments whatever is on the left by whatever is on the right.) We will also create a function to compute $1+2+\cdots+n$ for any integer $n>0$.

\begin{sageCell}
def mySum(n):
        s = 0
        for i in range(1,n+1):
                s += i
        return s

print(mySum(100))
\end{sageCell}

\section{While Loops}

The Python syntax for a while loop is as follows:

\begin{verbatim}
==============================
while condition:
        code
==============================
\end{verbatim}

In the above code, \verb|while| indicates the start of a while loop, \verb|condition| is the comparison that is made to determine if the while loop is finished, and \verb|code| is executed as long as \verb|condition| holds true.

Below is an example of how to compute $1+2+\cdots+100$ using a while loop.

\begin{sageCell}
n = 100
s = 0
i = 1

while i <= n:
        s += i
        i += 1

print(s)
\end{sageCell}

Note that for a while loop, if \verb|condition| is always true, then your algorithm will never finish running. Unlike a for loop, you must be sure to update the variables associated with the condition being checked at each iteration. In the example above, omitting the line \verb|i += 1| would create an infinite loop.

\section{For vs. While}

A common question when learning about these two types of loops is: Which one should I use? In many cases either will work just fine. For loops are preferred when the number of times the instructions are to be repeated is known. While loops are preferred when the number of times the instructions are to be repeated is unknown.

For example, we have already seen that computing $1+2+\cdots+n$ for some integer $n>0$ can be done with a for loop using just a few lines of code. Consider the following related question: What is the smallest integer $m$ such that $1+2+\cdots+m\geq 1000$?

In this case, a for loop may not be as helpful as a while loop since we do not know how many times to iterate. Using a while loop, we can track the value of the sum after each step. We can use the value of the sum as the condition that determines whether or not the while loop continues to iterate. See the code below.

\begin{sageCell}
s = 0
i = 1

while s < 1000:
        s += i
        i += 1
print(i-1)  # Why do we print i-1 and not i?
\end{sageCell}

\section{Problems}

The goal of these problems is to convert several of the flowchart solutions to questions in the Flowcharts With Operators to Python functions.

\begin{question}
	Define a function called \verb|fact| that computes $n!$ for any nonnegative integer $n$. (See question 3 from the Flowcharts With Operators section.)
	\begin{hint}
	Consider using \verb|range(1,n+1)| in your for loop and the \verb|*=| operator for updating the value of the factorial variable. Note that the second hint for this problem is the solution.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def fact(n):
        fac = 1
        for i in range(1,n+1):
                fac *= i
        return fac
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
	Define a function called \verb|div| that computes the number of divisors of $n$ for any integer $n>0$. (See question 5 from the Flowcharts With Operators section.)
	\begin{hint}
	Consider using a for loop since the divisors of $n$ are in the interval $[1,n]$. Note that the second hint for this problem is the solution.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def div(n):
        divCount = 0
        for i in range(1,n+1):
                if n%i == 0:
                        divCount += 1
        return divCount
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
	Define a function called \verb|prime| that determines if a positive integer $n$ is a prime number. If $n$ is a prime number, the output should be the boolean value \verb|True|, with \verb|False| otherwise. (See question 7 from the Flowcharts With Operators section.)
	\begin{hint}
	Consider using the previous function to help you determine if $n$ is prime. There are two ways to do this:
	\begin{itemize}
	\item Once you define a function, it can be used inside the definition of another.
	\item You can define a function inside of another. But beware of scope issues in this case.
	\end{itemize}
	Note that the second hint for this problem contains both solutions described above.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def div(n):
        divCount = 0
        for i in range(1,n+1):
                if n%i == 0:
                        divCount += 1
        return divCount
def prime(n):
        if div(n) == 2:
                p = True
        else:
                p = False
        return p
\end{sageCell}
\begin{sageCell}
def prime(n):
        def div(n):
                divCount = 0
                for i in range(1,n+1):
                        if n%i == 0:
                                divCount += 1
                return divCount

        if div(n) == 2:
                p = True
        else:
                p = False
        return p
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
	Define a function called \verb|pos_floor| that computes the floor of a real number $x\geq 0$. The floor of a real number $x$ is the largest integer less than or equal to $x$, in other words, rounding $x$ down. (See question 8 from the Flowcharts With Operators section.)
	\begin{hint}
	Consider using a while loop. Note that the second hint for this problem is the solution.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def pos_floor(x):
        n = 0
        while n <= x:
                n += 1
        return n-1
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
	Define a function called \verb|neg_floor| that computes the floor of a real number $x<0$. (See question 9 from the Flowcharts With Operators section.)
	\begin{hint}
	This problem can be solved in a similar way to the previous one. Note that the second hint for this problem is the solution.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def neg_floor(x):
        n = 0
        while n < -x:
                n += 1
        return -n
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
	Define a function called \verb|my_floor| that computes the floor of any real number $x$. (See question 10 from the Flowcharts With Operators section.)
	\begin{hint}
	Combine the solutions to the previous two problems. Note that the second hint for this problem is the solution.
	\end{hint}
	\begin{hint}
\begin{sageCell}
def pos_floor(x):
        n = 0
        while n <= x:
                n += 1
        return n-1
def neg_floor(x):
        n = 0
        while n < -x:
                n += 1
        return -n
def my_floor(x):
        if x >= 0:
                flr = pos_floor(x)
        else:
                flr = neg_floor(x)
        return flr
\end{sageCell}
	\end{hint}
\end{question}

\section{Workspace}

\begin{sageCell}
# Use this cell to solve the above questions.
\end{sageCell}

\end{document}
