\documentclass{ximera}  
\title{Lists}  
\begin{document}  
\begin{abstract}  
We give a brief discussion to lists, indexing, and slicing.
\end{abstract}  
\maketitle

\section{Lists}

A list is an ordered sequence of comma separated values. Members of a list are often referred to as elements and are assigned a position, known as their index, on the list. Lists are zero-indexed, meaning that the first element in the list is indexed by 0, the second by 1, and so on.

Below we give an example of a list and how to reference various elements in the list.

\begin{verbatim}
==============================

a = [1, 3, 6, -1, 5]

==============================
\end{verbatim}

The variable \verb|a| above is a list containing five elements. The first element on the list can be referenced by \verb|a[0]|, the second by \verb|a[1]|, etc. In Python, one can use negative numbers as well to go backwards, making the last element \verb|a[-1]|, the penultimate \verb|a[-2]|, etc. Note that to create a list, the sequence of comma separated values must be enclosed by square brackets \verb|[ ]|. Parentheses and curly braces are for different, but related, data types that we will avoid for now.

Lists are useful in that they can hold different data types and one can iterate over them. It is this second property that will be explained further and used extensively in the next section. 

\section{Functions and Operators}

Python comes equipped with functions and operators designed to work in a particular way when applied to lists. Here are some common examples:

\begin{itemize}
	\item Concatenation (\verb|+|) - This operator concatenates lists, so \verb|[1,2]+[3,4]| would yield \verb|[1,2,3,4]|.
	\item Multiplication (\verb|*|) - This operator requires a list and integer and is best illustrated via an example. We have that \verb|[1,2] * 3| yields \verb|[1,2,1,2,1,2]|.
	\item Length (\verb|len|) - This function computes the length of a list, so \verb|len([1,3])| yields \verb|2|.
	\item Membership (\verb|in|) - This keyword can be used to determine if a particular value is a member of a list. For example, \verb|2 in [1,2,3]| is \verb|True|.
\end{itemize}

\section{Slicing}

List slicing is a handy way of obtaining a portion of a list using the following notation. Given a list \verb|a|, \verb|a[start:stop]| will give all of the elements of \verb|a| starting with index \verb|start| and ending with index \verb|stop-1|. If either \verb|start| or \verb|stop| are omitted, then the longest list satisfying the contraints is given. Also, a third optional argument can be given to specify the step size taken. We illustrate various ways to slice a list below.

\begin{sageCell}
myList = ['cake', 'pie', 'ice cream', 'donut', 'brownie', 'fruit']

print(myList[2:5])  # Contains elements with index 2,3,4.
print(myList[:3])   # Contains elements with index 0,1,2.
print(myList[1:])   # Contains elements with index >= 1.
print(myList[2::2]) # Contains every other element with index >= 2.
print(myList[::-1]) # Reverses the list.
\end{sageCell}

\section{Caution}

Consider the following example:

\begin{sageCell}
a = [1,2,3]

b = a

b[1] = -7

print(a) # What do you expect to see here?
print(b)
\end{sageCell}

Note that \verb|b=a| does not create a new list \verb|b| with the same elements as \verb|a|. Rather, both are references to the same object. Any changes to \verb|b| will therefore be reflected by \verb|a| as well. To create a duplicate copy of a list that avoids this issue, use \verb|b = a[:]|.

\section{Problems}

Note that for the questions below, the hint contains the solution.

\begin{question}
Consider the list \verb|a = [4,5,2,1,6]|. Use list slicing to obtain the list \verb|2,1,6|.
	\begin{hint}
\begin{sageCell}
a = [4,5,2,1,6]
a[2:]
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
Write a Python function \verb|longer| that takes in two lists \verb|a| and \verb|b| as inputs and returns the length of the longer list.
	\begin{hint}
\begin{sageCell}
def longer(a,b):
        if len(a) >= len(b):
                long = len(a)
        else:
        long = len(b)
        return long
\end{sageCell}
	\end{hint}
\end{question}

\begin{question}
A list is a palindrome if it is the same forwards and backwards, like \verb|[1,2,3,4,3,2,1]|. Write a Python function \verb|pal| that takes in a list \verb|a| as an input and returns \verb|True| if the list is a palindrome and \verb|False| otherwise.
	\begin{hint}
\begin{sageCell}
def pal(a):
        return a == a[::-1]
\end{sageCell}
	\end{hint}
\end{question}

\section{Workspace}

\begin{sageCell}
# Use this cell to solve the above questions.
\end{sageCell}

\end{document}
