\documentclass{ximera}  
\title{Matplotlib}  
\begin{document}  
\begin{abstract}  
We give an introduction to plotting using Matplotlib.
\end{abstract}  
\maketitle

\section{Matplotlib}

The \link[Matplotlib]{https://matplotlib.org/} library is a plotting library that is commonly used for producing publication quality figures of all kinds. Matplotlib has many features (too many to cover here). The easiest way to create a plot is to use the example gallery found on the Matplotlib website. Find the type of plot and features that you want and use the code sample provided.

To illustrate the kinds of plots that we will generate, we begin with an example. The code below produces a plot of $f(x)=5x^2-400$ and $g(x)=x^3-x+1$ on the interval $[-10,10]$. 

\begin{sageCell}
import numpy as np
import matplotlib.pyplot as plt

x = np.linspace(-10,10,100)
y1 = 5*x**2-400
y2 = x**3-x+1
plt.plot(x,y1,color='blue',label='$f(x)$')
plt.plot(x,y2,color='red',label='$g(x)$')
plt.xlabel('$x$')
plt.ylabel('$y$')
plt.legend()
plt.title('A graph of $f(x)$ and $g(x)$')
plt.show()
\end{sageCell}



\begin{itemize}
	\item \verb|import matplotlib.pyplot as plt| - this imports the appropriate Matplotlib functions with the commonly used alias \verb|plt|
	\item \verb|np.linspace(-10,10,100)| - this produces an array containing 100 evenly spaced points from -10 to 10
	\item \verb|plt.plot| - this adds the data points to the image, additional arguments specify their color and add a label for the plot legend (use \verb|plt.scatter| to draw points without connecting them)
	\item \verb|plt.xlabel|, \verb|plt.ylabel| - this adds axes labels
	\item \verb|plt.legend| - this adds a legend to the graph
	\item \verb|plt.title| - this adds a title to the graph
	\item \verb|plt.show| - this displays the graph
\end{itemize}

\section{Problems}

Note that for the questions below, the hint contains the solution.

\begin{question}
\end{question}

\section{Workspace}

\begin{sageCell}
# Use this cell to solve the above questions.
\end{sageCell}

\end{document}
