\documentclass{ximera}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm, text centered, fill=red!50]
\tikzstyle{io} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum height=1cm, text centered, fill=blue!30]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, text width=3cm, fill=orange!50]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, text width=2cm, fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\title{Newton's Method (optional)}
\begin{document}
\begin{abstract}
We give an introduction to finding roots using Newton's Method.
\end{abstract}
\maketitle

\section{Linear Approximations}

In a calculus course, it is shown that the best linear approximation to a differentiable function $f(x)$ near $x=a$ is given by $L_a(x)=f'(a)(x-a)+f(a)$. We will make use of this general idea to try to approximate solutions of equations of the form $f(x)=0$.

The Desmos interactive graph below gives an example of a linear approximation to a differentiable function.

\desmos{adu2uowy3n}{800}{600}

\section{Newton's Method}

Suppose that $f(x)$ is a differentiable function and that we would like to solve the equation $f(x)=0$. Using the IVT, or a graphical method, we might have an initial guess for a solution that we will call $x_0$. We are interested in finding a value $r$ such that $f(r)=0$. We know that if $r$ and $x_0$ are ``close" to each other, then $$f(r)\approx f'(x_0)(r-x_0)+f(x_0).$$ Setting the right-hand side above equal to 0, solving for $r$ in terms of $x_0$, and renaming $r$ as $x_1$ gives that $$x_1=x_0-\frac{f(x_0)}{f'(x_0)}.$$ Iterating this process gives that $$x_n=x_{n-1}-\frac{f(x_{n-1})}{f'(x_{n-1})}$$ for $n\geq 1$. We can iterate the above procedure until $|f(x_n)|$ becomes as small as we like, at which point the claim is that $x_n$ is an approximate solution to the equation.

The Desmos interactive graph below gives an example of Newton's method iterated twice.

\desmos{zokv8syqrv}{800}{600}

Before we go over the algorithm for Newton's Method, we must note a disadvantage of this method. Unlike the Binomial Method, Newton's Method may not converge (try applying the method to $f(x)=x^{1/3}$). We must therefore specify a maximum number of iterations we are willing to compute to ensure that the algorithm terminates.

We can summarize Newton's Method using the following flowchart (see Figure 1). Here we assume that $f(x)$ is differentiable, that $f'(x)\neq 0$, that $x_0$ is close to a solution of $f(x)=0$, that $n$ is the maximum number of iterations we are willing to compute, and that $err>0$.

\begin{center}
    \begin{tikzpicture}
    \node [startstop] at (0,0) (start) {Start};
    \node [io] at (0,-2) (io1) {$f,f',x_0,n,err$};
    \node [process] at (0,-4) (process1) {$x:=x_0$\\$i:=0$};
    \node [decision] at (0,-7) (decision1) {$|f(x)|>err$\\$i\leq n$};
    \node [decision] at (-5,-7) (decision2) {$i>n$};
    \node [process] at (0,-10) (process2) {$x:=x-\displaystyle\frac{f(x)}{f'(x)}$\\$i:=i+1$};
    \node [io] at (-5,-10) (io2) {Output $x$};
    \node [io] at (-10,-7) (io3) {Output ``Error"};
    \node [startstop] at (-10,-10) (stop) {Stop};
    \draw [arrow] (start) -- (io1);
    \draw [arrow] (io1) -- (process1);
    \draw [arrow] (process1) -- (decision1);
    \draw [arrow] (decision1) -- node[anchor=east] {yes} (process2);
    \draw [arrow] (decision1) -- node[anchor=south] {no} (decision2);
    \draw [arrow] (decision2) -- node[anchor=east] {no} (io2);
    \draw [arrow] (decision2) -- node[anchor=south] {yes} (io3);
    \draw [arrow] (process2) -- (3,-10) |- (decision1);
    \draw [arrow] (io2) -- (stop);
    \draw [arrow] (io3) -- (stop);
    \end{tikzpicture}
\end{center}
\begin{center}
	Figure 1: An algorithm for approximating roots using Newton's Method.
\end{center}

Converting this flowchart to code gives the following:

\begin{verbatim}
==============================
def newton(f,df,x0,n,error):
        x = x0
        i = 0
        while abs(f(x)) > error and i <= n:
                x = x - f(x)/df(x)
                i += 1
        if i > n:
                return "Exceeded iterations"
        else:
                return x
==============================
\end{verbatim}

While Newton's Method may fail to converge, one advantage of this method is that it converges to a solution faster than the Binomial Method in many cases.

\section{Problems}

All answers in this section have been rounded to the nearest thousandth, the error for the output has been set to $0.0001$, and the number of iterations have been capped at 10.

\begin{question}
	Use Newton's Method to estimate the solution to $f(x)=0$ where $f(x)=x^3-5$. $\answer{1.710}$
\end{question}

\begin{question}
	Use the Newton's Method to give an approximation of $\sqrt{3}$. $\answer{1.732}$
\end{question}

\begin{question}
	Use Newton's Method to give an approximation of the solutions to the equation $x\ln^2{x}=1/2$. Enter your solutions in incresing order. $\answer{0.073}$ $\answer{0.226}$ $\answer{1.716}$
	\begin{hint}
	Different initial values will converge to different solutions. Use a graphical method to determine different initial values.
	\end{hint}
\end{question}

\section{Workspace}

\begin{sageCell}
# Use this cell to solve the above questions.
\end{sageCell}

\end{document}
